The illusory promise of the Aligned Rank Transform

### 2025-07-07 text/30-assumptions.qmd
minor change

## Revisiting ART's distributional assumptions {#assumptions}
Although ART is often regarded as a nonparametric method, its alignment mechanism relies on a set of strict assumptions.
For a two-way experimental design with independent observation, @higgins:1990 assume the following model for responses:

@@

<!--
Thus, the critical question we must address is: how does ART perform when its core assumptions are violated?
-->

@@

However, a question remains: Is ART’s failure in handling such models due solely to heteroscedasticity issues? For example, consider the two time distributions in @fig-homoscedastic-model, which --- despite having different shapes --- share the same standard deviation. Would ART perform correctly in this case? While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Naturally, as the shapes become more similar, the problem diminishes. In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.

@@

>>>>

## Revisiting ART's distributional assumptions {#assumptions}
ART is often regarded as a nonparametric method, but its alignment mechanism relies on a set of strict assumptions. For a two-way experimental design with independent observation, @higgins:1990 assume the following model for the responses:

@@

@@

However, a question remains: Is ART’s failure in handling such models due solely to heteroscedasticity issues? For example, consider the two time distributions in @fig-homoscedastic-model, which --- despite having different shapes --- share the same standard deviation. Would ART perform correctly in this case?

@@

While we do not directly address this question in the main paper, we include additional results in the [appendix](appendix.html) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Naturally, as the shapes become more similar, the problem diminishes. In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.
